Positivity-Preserving Adaptive Runge-Kutta Methods

### 2020-02-18 latex/overview.tex
Changes in Introduction, Corrections
overview.tex
Minor corrections

Bib_RK_optim.bib:
New sources

Therefore, it is interesting to know how to use RKMs that preserve positifity.
For RKMs some theoretical limitations are known. One is that any unconditionally positive RKM has an order $\leq 1$. \cite{hundsdorfer_numerical_2003,bolley_conservation_1978,horvath_positivity_1998}. The only well know RKM that is unconditionally positive is the backward euler method, which is obliviously of first order. 
This causes restrictions on the step size for any RKM with a higher order.

There are multiple approaches to get positivity preserving numerical schemes.
One method is the Strong Stability Preserving Runge Kutta (SSPRK). 
The idea is to construct a RKM that can be written as convex combination of euler steps. 
By this the stability properties of the forward euler step are carried over to the SSPRK.
The same happens to the positivity properties. By this the positivity of the method is ensured for time steps $\dt$ with $\dt \leq R \dt_{FE}$, where $R$ depends on the used SSPRK and $\dt_{FE}$ is the forward euler time step \cite{gottlieb_strong_2011}. 
The Modified Patanka Runge Kutta (MPRK) are 
based on the Patanka RKM. In these destruction Terms are scaled by the new solution. This preserves positivity but violates the mass conservation. The MPRK overcome this disadvantage by also scaling the production terms. This is done by solving a linear equation system \cite{kopecz_comparison_2019}.
An approach to ensure positivity in a practical application is shown in \cite{shampine_non-negative_2005}. Ensuring the positivity is done by redefining the ODE outside the domain. Another technique used are event finding methods. This approach is implemented in the Matlab ODE Suite.
Unconditionally positive method of order higher than 1 are the diagonally split Runge-Kutta (DSRK) methods \cite{horvath_positivity_1998}. These are no longer part of the general linear methods. This allows them to be positive and higher order. (TODO: two sentences here)

@@

A common way to calculate two different solutions with one set of stage values is the use of embedded methods for error approximation. 
The idea of adapting the weights $b$ after calculating the stage values is used in \cite{ketcheson_spatially_2013}
In this case it is used to adapt the properties of the time integrator for a method of lines simulation on a PDE. With this approach, it is possible to have different properties of the RKM at different parts of the domain.

In this paper, an adaption of the $b$ is done by solving a Linear programming (LP)-Problem.
(TODO: Description LP-Problem)

The paper unfolds as follows. In section\,\ref{sec:main_idea} the main idea is explained. How the method can be written as LP-Problem is explained in section\,\ref{sec:LP-Problem}.
In section\,\ref{sec:integration} it is described how the new approach can be used with different RKM, how it can be integrated in a step size control and how the Region of absolute stability can be approximated.  
In section\,\ref{sec:imple} further details that were used when implementing the algorithm are described.
In section\,\ref{sec:Numeric_Results} Numerical results are given for multiple testproblems.
A conclusion is given in section\,\ref{sec:conclusion}

@@

In the following section, it is explained how the approach can be written as a LP-Problem. 
For this we have to consider which $b$ to choose. 
Generally, any $b \in \R$ can be used. 
A first requirement would be that the new solution is positive.
We can further reduce this space by enforcing the Order Conditions. 
By incorporating the Order Conditions, the space of $b$ is limited to a subspace of $\R$. %(Note: not linear because 0 is not in subspace)
Finally, we have to give a objective function to get the optimal $b$.

@@

\begin{equation}
u'(t) = L u(t) \qquad L = \left[\begin{matrix}- a & 1\\a & -1\end{matrix}\right] \quad a =5
\end{equation}
and the first order RKM with two stages
\begin{align}
\begin{array}{c|cc}
0 &  & \\
\frac{1}{2} & \frac{1}{2} & \\
\hline
 & \frac{1}{2} & \frac{1}{2}\\
\end{array}
\end{align}
If simulated with $\dt = 1$ the stage values are 
\begin{equation}
f^1 = \left(\begin{matrix}-5\\5\end{matrix}\right) \qquad f^2 = \left(\begin{matrix}10\\-10\end{matrix}\right)
\end{equation}
All $b$ that comply with the order Condition for the first order can be expressed as
\begin{equation}
b= \left(\begin{matrix}\frac{1}{2}\\ \frac{1}{2}\end{matrix}\right) + \alpha \left(\begin{matrix}1\\ -1\end{matrix}\right)
\end{equation}
With this we have one degree of freedom for the choice of $b$. This is done by the parameter $\alpha$.
If the general expression for $b$ is inserted in \eqref{eq:Combination} the general solution is 
\begin{equation}
u^{1} = u^0 + \dt  \left[f^1,f^2\right] b = \left(\begin{matrix}\frac{5}{2}\\- \frac{5}{2}\end{matrix}\right) \alpha  \left(\begin{matrix}-15\\15\end{matrix}\right)
\end{equation}
The solution for the original RKM would be $u^{1} = \left(\frac{5}{2} ,- \frac{5}{2}\right)^T$ which contains a negative value. But by changing the parameter $\alpha$ and therefore changing the $b$ a positive solution can be reached.
By a suitable choice of $\alpha \in \left[\frac{1}{6},\frac{7}{30}\right]$ any $u$ that complies with mass conservation and positivity can be reached. 
By adding additional constraints on the $b$ the choice of $\alpha$ can be narrowed down. 
An objective function is also needed to make the choice unique. This should be designed in a way to prefer $b$ that are close to the original weights $b_{orig}$

@@

The Order Conditions for a RKM are a set of Equations depending on $a_{jl}$, $c_1,\cdots,c_s$ and $b_1,\cdots,b_s$. In these Equations $b_1,\cdots,b_s$ only appear linearly \cite{hairer_runge-kutta_1993}.
If $a_{jl}$ and $c_1,\cdots,c_s$ are already known, the Order Conditions for the Order $p$ are an linear equations system and can be written as $Q b = r$.

The vector $b \in \R^s$ contains the weights and $Q$ is an $n \times s$ matrix, where $g$ is the number of Order Conditions. The vector $r$ contains the right hand side of the Order Conditions.

If the $A$ of a preexisting RK-method is used, there is at least one solution for the linear equation system.
In order to get an optimization problem, the equation system has to be underdetermined.
This implies that $\mathrm{rank}(Q) < s$. 
Because the Quadrature Conditions are linear independent, there are at least $p$ independent vectors in $Q$. This leads to the condition $p < s$.
This implies that the number of stages has to be higher as the wanted order.

@@

Both approaches are shown in figure\,\ref{fig:b_space}.

@@

When directly adapting the $b$ a new $b$ that complies with the Order Conditions is chosen.   
The weights $b = b_{orig} + \Delta b$ consist of the weights of the original method $b_{orig}$ and an adaption $\Delta b$. The adaption is $\Delta b \in \mathrm{ker}(Q)$. 
In Figure\,\ref{fig:b_direct} this is drawn. The red line represents the subspace of $b$ that satisfy the Order Conditions. Both $b_{orig}$ and $b$ are in this subspace.

@@

\subsubsection{Constraints}
The Direct Adaptation approach leads to two set of constraints on the $b$.
The first set of constraints is the positivity and can be written as 
\begin{equation}
u_{n+1}=u_n+\dt K \vec{b} \geq 0
\end{equation}
The second set of constraints are the Order Conditions
\begin{equation}
Qb=r
\end{equation}

@@

A reasonable property of the objective function would be that for the LP-Problem only constrained by the order condition the optimal solution is $b = b_{orig}$. 
This can be achieved by $argmin(f_{optim}(b)) = b_{orig}$. 
In practice, this also means that there is no need to solve solve the optimization Problem if no constraints are violated because the solution of this problem is already known. This is advantageous for performance reasons.
The different approaches have been tested on a linear test problem from \cite{kopecz_unconditionally_2018}. The optimization problem was used at every step.
The magnitude of $b$ can get large if the Order Conditions are used as objective function.
Large values of b can lead to numerical problems. Because $\sum_{i  = 1}^s b_i = 1$, an increase of $||b||$ leads to a combination of negative and positive values. Because the computation is done with a finite precision (and usually in floating point) large $||b||$ can lead to absorption effects.    
By changing the $b$ the stability function is altered. This can lead to oscillations if the poles of the RHS are no longer in the stability region of the resulting method.
For these reasons $||b-b_{orig}||$ seems like a good choice. The 1-Norm can be implement the with Linear Programming using slack variables. We choose

@@

If the original step $\dt$ was too large and there is no suitable $b$, we would like to calculate a $u^{n+1}$ using the stage values instead of rejecting the step. For this the time step is reduced after calculation the stage values. This can be done using the Order Conditions for dense Output. These can be found in \cite{hairer_runge-kutta_1993}.
We introduce $\theta \in (0,1]$. The new step taken has the length $\theta \dt$ whereas $\dt$ is the time step used when calculating the stage values.
To calculate a reduced step the Order Conditions have to be adapted and a new $b_{orig}$ for the objective function (and more important the Error detection afterwards) has to be generated. 
(TODO: the $b_{orig}$ is still a open question)

@@

\subsubsection{Constraints}
At first a set of embedded methods of the RKM $b_1,\cdots,b_d$ are needed. 
If all the used $b$ compile with the order conditions the new method is also of the same Order.
This is straightforward using
$$ Q b = Q (\chi b_{\alpha} + (1-\chi) b_{\beta}) = \chi Q  b_{\alpha} + (1-\chi) Q b_{\beta}) = \chi r + (1-\chi) r = r$$
$b_{\alpha}$ and $b_{\beta}$ are the methods that are combined to $b$ using the parameter $\chi$. The Matrix $Q$ and the vector $r$ are the Order Conditions as defined in Section\,\ref{sec:OrderCond}.

@@

\subsubsection{Construction of Embedded Methods}
(TODO: have a look on which of the ideas are still valid, maybe rewrite section completely)

For methods with $s \geq p$ it is useful to have some degrees of freedom for the highest order. For this at least one $b$ that complies with the order Condition of the highest order and $b \neq b_{orig}$ has been introduced.
These additional $b$ can be constructed by calculating a Span $S$ of the Kernel $\mathrm{ker} (Q)$ of the Order Condition Matrix $Q$. For every vector $s \in S$ one set of weights with $b = b_{orig} + a s$ and one with $b = b_{orig} - a s$ is added. The $a$ is a parameter that determines how big a change is allowed.

As a first order method, a $b$ is added that corresponds to a series of backward euler steps. These are known to be yield to a positive result.

An important property of the embedded methods is that they should show the desired stability characteristics.

A Reduction of the step size suing the order Conditions for dense output would also be possible. For this a set of embedded methods $b_1(\Theta),\cdots,b_d(\Theta)$ has to be constructed. This was not done in this research.

(Note: Doing dense Output here is a bit complicated, but it would not help ensure positivity only (maybe) increase exactness.  is it worth exploring this or is it better to just leave this open for further research it may be also interesting regardless the study of positivity)
Everything has downsides...

@@

\subsection{Error detection and Approximation}
When changing the weights the solution $u^{n+1}$ is changed. 
By incorporating the Order Conditions for a certain Order the space of $b$ that can be choose is limited to a linear Subspace. 
Now it is important to know if the new solution $u^{n+1}$ still resembles the solution of the ODE $u(t_{n+1})$. 
This can be seen in example in section\,\ref{sec:example_lin}.
By a suitable choice of $\alpha \in [\frac{1}{6},\frac{7}{30}]$ any $u$ that complies with mass conservation and positifity can be reached.

@@

The derivatives of $u(t)$ can be solely expressed with the partial derivatives of the RHS $f(t,u)$, whereas the coefficients $v_1,v_2,\cdots$ also depend on the RKM. 
In this case the RKM is fixed apart from the weights $b$. 
Because of this $v_1,v_2,\cdots$ are functions of $b$.
Because the $b$ are chosen to ensure the Order Conditions the coefficients $v_1,\cdots,v_p$ are fixed and equal the according derivatives. 
The only remaining degrees of freedom are the coefficients $v_{p+1},v_{p+2},\cdots$.
These can be adapted by changing the $b$.

@@

The stability function is defined as 
\begin{equation}\label{eq:stab_func}
R(z) = 1 + zb^T(I - zA)^{-1}e
\end{equation}
We can define a function $P(z)$
\begin{equation}
P(z) = R(z)-1 = z b^T (I - zA)^{-1}
\end{equation}
These Polynomials are $\in \mathcal{L}^\infty$. (Note: for all explicit methods this is always true, for implicit methods we have to exclude the poles $\{z | \det(I - zA) = 0\}$ or  $\{z | \det(I - zA) > \epsilon \}$ )
We define $\mathcal{P}$ as all $P(z)$ for the class of $RKM_A$. 
This is a vector space.
Proof:
\begin{multline}
\alpha P_a(z) + \gamma P_c(z) =  \alpha z a^T (I - zA)^{-1} + \gamma z c^T (I - zA)^{-1} =\\
z (\alpha a+ \gamma c)^T (I - zA)^{-1} = P_{\alpha a + \gamma c}(z) \in \mathcal{P}
\end{multline}

The $P(z)$ is a linear combination of the Polynomials
\begin{subequations}
\begin{align}
p_1(z) &= z e_1^T (I - zA)^{-1}\\
\vdots \\
p_s(z) &= z e_s^T (I - zA)^{-1}
\end{align}
\end{subequations}
whereas $e_1,\cdots,e_s$ is the standard basis of $\R^s$.
These span $\mathcal{P}$. If $p_{e_1}(z),\cdots,p_{e_s}(z)$ are linearly independent $p_{e_1}(z),\cdots,p_{e_s}(z)$ are a basis of $\mathcal{P}$.

@@

\subsubsection{Stability of convex combination}
If the new $b$ is chosen as a convex combination of $b$ it is easy to prove properties of the stability region.
The stability function $R_{\chi a + (1-\chi) c}(z)$ of a convex combination can be calculated by the weighted sum of the Stability functions $R_a(z)$ and $R_c(z)$

Proof:
\begin{multline}
R_{\chi a + (1-\chi) c}(z) = R(z) = 1 + (\chi a + (1-\chi) c)^Tz(I - zA)^{-1}e \\
= \chi + (1-\chi) + \chi a ^Tz(I - zA)^{-1}e + (1-\chi) c^Tz(I - zA)^{-1}e \\
= \chi R_a(z) + (1-\chi) R_c(z)
\end{multline}

@@

For most methods, a small change of $b$ does not alter the stability function dramatically.
(TODO: I am missing a piece here from ‘the derivative is finite’ to ‘is simmilar’)

@@

The stability function depending on the $b$ is expressed 
When adapting the $b$ we cannot change them independently but only in ways that comply with the order conditions
$m$ is the number of dimensions of the Kernel of the Order Condition $m = \mathrm{dim}\{\mathrm{Ker} (Q) \}$
The $b_1,\cdots,b_m$  are the Basis vectors of $\mathrm{Ker} (Q)$. With an coordinate Transformation we get the coordinate vector $\hat{b}$ with the Matrix $B = [b_1,\cdots,b_m]$ and $b = B \hat{b}$.

@@

\begin{equation}\label{eq:gen_stabilityf}
R(\hat{b},z) = 1 +  (b_{orig} +B \hat{b})^T z(I - zA)^{-1}
\end{equation}
The border is defined by the implicit function 
\begin{equation}\label{eq:border}
|R(\hat{b},u+iv)|^2 -1 = 0
\end{equation}
This function can be viewed either as $\R^m \times  \CN \rightarrow \R$ or $\R^m \times  \R^2 \rightarrow \R$.
We are only concerned with the expansion or contraction. Because of this we are only interested in the change perpendicular to the Border of the stability region.

@@

The derivative in respect to $\hat{b}_j$ is 
$ \frac{\partial f}{\partial \hat{b}_j}(\hat{b},a(\hat{b})) \Big|_{\hat{b}=0}$
this simplifies to 
\begin{multline}\label{eq:derivative_to_b}
 \frac{\partial f}{\partial \hat{b}_j}(w,a(w)) \Big|_{\hat{b}=0} = 
 \frac{\partial }{\partial \hat{b}}(R(\hat{b},z)R(\hat{b},z^*)) \\
 = (B\hat{b})^T (I-zA)^{-1} R_{b_{orig}}(z^*) + (B\hat{b})^T (I-z^*A)^{-1} R_{b_{orig}}(z)
\end{multline}
using the fact that $z_0$ is on the border of the stability region.
The derivative $(B\hat{b})^T (I-zA)^{-1} R_{b_{orig}}(z^*) (B\hat{b})^T (I-z^*A)^{-1} R_{b_{orig}}(z)$ is known to be finite for explicit methods, because it is a polynomial and $|z_0| < \infty$
The same process can be done for all points on the border of the stability region. 
As long as $\nabla|R(\hat{b},u+iv)|^2 \neq 0 \forall \hat{b} < \epsilon  \forall z$ the derivative $\frac{\partial a}{\partial \hat{b}_j} (\hat{b})$ is finite. This means that a small change the $b$ only leads to a small change of the stability region.

@@

\begin{align}
 u_i^{n+1} &\geq 0   \;   \forall {i \in \{1, \cdots,m \}}  \\
 u_i^n + h \sum_{j=0}^{s-1} b_j k_i^j  &\geq 0   \;   \forall {i \in \{1,\cdots,m \}}  
\end{align}
has to be fulfilled.
This infers $m$ positivity constraints to the optimization problem. These can be written as

\begin{equation}
u_i + \dt K  b \geq 0     
\end{equation}
where $K = \big[k^1 , \cdots k^{s-1}\big]$.

This causes unnecessary many inequality constraints on the LP-Problem. To simplify the problem, one want to reduce the number of positivity constraints by ignoring it for ODEs where there is no issue with positivity.
The new positivity constraints can be formulated as
\begin{equation}
u_i^n + \Delta b \sum_{j=0}^{s-1} b_j k_i^j  \geq 0   \;   \forall {i \in h \subseteq \{1,\cdot,m \}} 
\end{equation}
For this a set $h \subseteq \{1,\cdots,m \}$ that include all indecies that are needed for determining the $b$ has to be chosen.

>>>>

Therefore, it is interesting to know how to use RKMs that preserve positivity.
For RKMs some theoretical limitations are known. One is that any unconditionally positive RKM has an order $\leq 1$. \cite{hundsdorfer_numerical_2003,bolley_conservation_1978,horvath_positivity_1998}. The only well know RKM that is unconditionally positive is the backward euler method, which is obliviously of first order. 
This causes restrictions on the step size for any RKM with a higher order.

There are multiple approaches to get positivity preserving numerical schemes.
One method is the Strong Stability Preserving Runge Kutta (SSPRK). 
The idea is to construct a RKM that can be written as convex combination of euler steps. 
By this the stability properties of the forward euler step are carried over to the SSPRK.
The same happens to the positivity properties. By this the positivity of the method is ensured for time steps $\dt$ with $\dt \leq R \dt_{FE}$, where $R$ depends on the used SSPRK and $\dt_{FE}$ is the forward euler time step \cite{gottlieb_strong_2011}. 
The Modified Patanka Runge Kutta (MPRK) are 
based on the Patanka RKM. In these destruction Terms are scaled by the new solution. This preserves positivity but violates the mass conservation. The MPRK overcome this disadvantage by also scaling the production terms. This is done by solving a linear equation system \cite{kopecz_comparison_2019}.
An approach to ensure positivity in a practical application is shown in \cite{shampine_non-negative_2005}. Ensuring the positivity is done by redefining the ODE outside the domain. Another technique used are event finding methods. This approach is implemented in the Matlab ODE Suite.
Unconditionally positive method of order higher than 1 are the diagonally split Runge-Kutta (DSRK) methods \cite{horvath_positivity_1998}. These are no longer part of the general linear methods. This allows them to be positive and higher order.

@@

A common way to calculate two different solutions with one set of stage values is the use of embedded methods for error approximation. 
The idea of adapting the weights $b$ after calculating the stage values is used in \cite{ketcheson_spatially_2013}
In this case it is used to adapt the properties of the time integrator for a method of lines simulation on a PDE. With this approach, it is possible to have different properties of the RKM at different parts of the domain. 
Another method that adapts the $b$ are the relaxation Runge–Kutta (RRK) methods. 
In these the weights are scaled by a scalar relaxation parameter to conserve convex quantities. This can be used to conserve the energy or entropy \cite{ranocha_relaxation_2019,ketcheson_relaxation_2019}

In this paper, an adaption of the $b$ is done by solving a Linear programming (LP)-Problem.
These are convex optimization problems where all constraints and the objective function are all affine. These can incorporate both equality and inequality constraints. 
LP-Problems can be quite reliably solved with the Simplex method or with interior point methods. This can be done in reasonable time even for large problems. 
\cite{boyd_convex_2004}

The paper unfolds as follows. In section\,\ref{sec:main_idea} the main idea is explained. How the method can be written as LP-Problem is explained in section\,\ref{sec:LP-Problem}.
In section\,\ref{sec:integration} it is described how the new approach can be used with different RKM, how it can be integrated in a step size control and how the Region of absolute stability can be approximated.  
In section\,\ref{sec:imple} further details that were used when implementing the algorithm are described.
In section\,\ref{sec:Numeric_Results} Numerical results are given for multiple test problems.
A conclusion is given in section\,\ref{sec:conclusion}

@@

In the following section, it is explained how the approach can be written as a LP-Problem. 
For this we have to consider which $b$ to choose. 
Generally, any $b \in \R$ can be used. 
A first requirement would be that the new solution is positive.
We can further reduce this space by enforcing the order conditions. 
By incorporating the order conditions, the space of $b$ is limited to a subspace of $\R$. %(Note: not linear because 0 is not in subspace)
Finally, we have to give a objective function to get the optimal $b$.

@@

\begin{equation}
u'(t) = L u(t) \qquad L = \left[\begin{matrix}- a & 1\\a & -1\end{matrix}\right] \quad a =5
\end{equation}
and the first order RKM with two stages
\begin{align}
\begin{array}{c|cc}
0 &  & \\
\frac{1}{2} & \frac{1}{2} & \\
\hline
 & \frac{1}{2} & \frac{1}{2}\\
\end{array}
\end{align}
If simulated with $\dt = 1$ the stage derivatives are 
\begin{equation}
f^1 = \left(\begin{matrix}-5\\5\end{matrix}\right) \qquad f^2 = \left(\begin{matrix}10\\-10\end{matrix}\right)
\end{equation}
All $b$ that comply with the order Condition for the first order can be expressed as
\begin{equation}
b= \left(\begin{matrix}\frac{1}{2}\\ \frac{1}{2}\end{matrix}\right) + \alpha \left(\begin{matrix}1\\ -1\end{matrix}\right)
\end{equation}
With this we have one degree of freedom for the choice of $b$. This is done by the parameter $\alpha$.
If the general expression for $b$ is inserted in \eqref{eq:Combination} the general solution is 
\begin{equation}
u^{1} = u^0 + \dt  \left[f^1,f^2\right] b = \left(\begin{matrix}\frac{5}{2}\\- \frac{5}{2}\end{matrix}\right) \alpha  \left(\begin{matrix}-15\\15\end{matrix}\right)
\end{equation}
The solution for the original RKM would be $u^{1} = \left(\frac{5}{2} ,- \frac{5}{2}\right)^T$ which contains a negative value. But by changing the parameter $\alpha$ the weights $b$ and the solution is altered.
By a suitable choice of $\alpha \in \left[\frac{1}{6},\frac{7}{30}\right]$ any $u$ that complies with mass conservation and positivity can be reached. 
By adding additional constraints on the $b$ the choice of $\alpha$ can be narrowed down. 
An objective function is also needed to make the choice unique. This should be designed in a way to prefer $b$ that are close to the original weights $b_{orig}$

@@

The order conditions for a RKM are a set of Equations depending on $a_{jl}$, $c_1,\cdots,c_s$ and $b_1,\cdots,b_s$. In these Equations $b_1,\cdots,b_s$ only appear linearly \cite{hairer_runge-kutta_1993}.
If $a_{jl}$ and $c_1,\cdots,c_s$ are already known, the order conditions for the Order $p$ are an linear equations system and can be written as $Q b = r$.

The vector $b \in \R^s$ contains the weights and $Q$ is an $g \times s$ matrix, where $g$ is the number of order conditions. The vector $r$ contains the right hand side of the order conditions.

If the $A$ of a preexisting RK-method is used, there is at least one solution for the linear equation system.
In order to get an optimization problem, the equation system has to be underdetermined.
This implies that $\mathrm{rank}(Q) < s$. 
Because the Quadrature conditions are linear independent, there are at least $p$ independent vectors in $Q$. This leads to the condition $p < s$.
This implies that the number of stages has to be higher than the wanted order.

@@

Both approaches are illustrated in figure\,\ref{fig:b_space}.

@@

When directly adapting the $b$ a new $b$ that complies with the order conditions is chosen.   
The weights $b = b_{orig} + \Delta b$ consist of the weights of the original method $b_{orig}$ and an adaption $\Delta b$. The adaption is $\Delta b \in \mathrm{ker}(Q)$. 
In Figure\,\ref{fig:b_direct} this is drawn. The red line represents the subspace of $b$ that satisfy the order conditions. Both $b_{orig}$ and $b$ are in this subspace.

@@

\subsubsection{Constraints}
The Direct Adaptation approach leads to two set of constraints on the $b$.
The first set of constraints is the positivity and can be written as 
\begin{equation}
u_{n+1}=u_n+\dt K \vec{b} \geq 0
\end{equation}
The second set of constraints are the order conditions
\begin{equation}
Qb=r
\end{equation}

@@

A reasonable property of the objective function would be that for the LP-Problem only constrained by the order condition the optimal solution is $b = b_{orig}$. 
This can be achieved by $argmin(f_{optim}(b)) = b_{orig}$. 
In practice, this also means that there is no need to solve solve the optimization Problem if no constraints are violated because the solution of this problem is already known. This is advantageous for performance reasons.
The different approaches have been tested on a linear test problem from \cite{kopecz_unconditionally_2018}. The optimization problem was used at every step.
The magnitude of $b$ can get large if the order conditions are used as objective function.
Large values of b can lead to numerical problems. Because $\sum_{i  = 1}^s b_i = 1$, an increase of $||b||$ leads to a combination of negative and positive values. Because the computation is done with a finite precision (and usually in floating point) large $||b||$ can lead to absorption effects.    
By changing the $b$ the stability function is altered. This can lead to oscillations if the poles of the RHS are no longer in the stability region of the resulting method.
For these reasons $||b-b_{orig}||$ seems like a good choice. The 1-Norm can be implement the with Linear Programming using slack variables. We choose

@@

If the original step $\dt$ was too large and there is no suitable $b$, we would like to calculate a $u^{n+1}$ using the stage values instead of rejecting the step. For this the time step is reduced after calculation the stage values. This can be done using the order conditions for dense Output. These can be found in \cite{hairer_runge-kutta_1993}.
We introduce $\theta \in (0,1]$. The new step taken has the length $\theta \dt$ whereas $\dt$ is the time step used when calculating the stage values.
To calculate a reduced step the order conditions have to be adapted and a new $b_{orig}$ for the objective function (and more important the Error detection afterwards) has to be generated. 
(TODO: the $b_{orig}$ is still a open question)

@@

\subsubsection{Constraints}
At first a set of embedded methods of the RKM $b_1,\cdots,b_d$ are needed. 
If all the used $b$ compile with the order conditions the new method is also of the same Order.
This is straightforward using
$$ Q b = Q (\chi b_{\alpha} + (1-\chi) b_{\beta}) = \chi Q  b_{\alpha} + (1-\chi) Q b_{\beta}) = \chi r + (1-\chi) r = r$$
$b_{\alpha}$ and $b_{\beta}$ are the methods that are combined to $b$ using the parameter $\chi$. The Matrix $Q$ and the vector $r$ are the order conditions as defined in Section\,\ref{sec:OrderCond}.

@@

\subsubsection{Construction of Embedded Methods}

As a first order method, a $b$ is added that corresponds to a series of backward euler steps. These are known to be yield to a positive result for any step-size.
It is also possible to add embedded methods of higher order.
An important property of the embedded methods is that they should show the desired stability characteristics.

A Reduction of the step size suing the order conditions for dense output would also be possible. For this a set of embedded methods $b_1(\Theta),\cdots,b_d(\Theta)$ has to be constructed. This was not done in this research.

(Note: Doing dense Output here is a bit complicated, but it would not help ensure positivity only (maybe) increase exactness.  is it worth exploring this or is it better to just leave this open for further research it may be also interesting regardless the study of positivity)

@@

\subsection{Error detection and Approximation}
When changing the weights the solution $u^{n+1}$ is changed. 
By incorporating the order conditions for a certain order the space of $b$ that can be choose is limited to a linear Subspace. 
Now it is important to know if the new solution $u^{n+1}$ still resembles the solution of the ODE $u(t_{n+1})$. 
This can be seen in example in section\,\ref{sec:example_lin}.
By a suitable choice of $\alpha \in [\frac{1}{6},\frac{7}{30}]$ any $u$ that complies with mass conservation and positifity can be reached.

@@

The derivatives of $u(t)$ can be solely expressed with the partial derivatives of the RHS $f(t,u)$, whereas the coefficients $v_1,v_2,\cdots$ also depend on the RKM. 
In this case the RKM is fixed apart from the weights $b$. 
Therefore $v_1,v_2,\cdots$ are functions of $b$.
Because the $b$ are chosen to ensure the Order Conditions the coefficients $v_1,\cdots,v_p$ are fixed and equal the according derivatives. 
The only remaining degrees of freedom are the coefficients $v_{p+1},v_{p+2},\cdots$.
These can be adapted by changing the $b$.

@@

The stability function is defined as 
\begin{equation}\label{eq:stab_func}
R(z) = 1 + zb^T(I - zA)^{-1}e
\end{equation}
We can define a function $P(z)$
\begin{equation}
P(z) = R(z)-1 = z b^T (I - zA)^{-1}
\end{equation}
These rational functions are $\in \mathcal{L}^\infty$. (Note: for all explicit methods this is always true, for implicit methods we have to exclude the poles $\{z | \det(I - zA) = 0\}$ or  $\{z | \det(I - zA) > \epsilon \}$ )
We define $\mathcal{P}$ as all $P(z)$ for the class of $RKM_A$. 
This is a vector space.
Proof:
\begin{multline}
\alpha P_a(z) + \gamma P_c(z) =  \alpha z a^T (I - zA)^{-1} + \gamma z c^T (I - zA)^{-1} =\\
z (\alpha a+ \gamma c)^T (I - zA)^{-1} = P_{\alpha a + \gamma c}(z) \in \mathcal{P}
\end{multline}

The $P(z)$ is a linear combination of the rational functions
\begin{subequations}
\begin{align}
p_1(z) &= z e_1^T (I - zA)^{-1}\\
\vdots \\
p_s(z) &= z e_s^T (I - zA)^{-1}
\end{align}
\end{subequations}
whereas $e_1,\cdots,e_s$ is the standard basis of $\R^s$.
These span $\mathcal{P}$. If $p_{e_1}(z),\cdots,p_{e_s}(z)$ are linearly independent $p_{e_1}(z),\cdots,p_{e_s}(z)$ are a basis of $\mathcal{P}$.

@@

\subsubsection{Stability of convex combination}
If the new $b$ is chosen as a convex combination of $b$ it is easy to prove properties of the stability region.
The stability function $R_{\chi b_a + (1-\chi) b_b}(z)$ of a convex combination can be calculated by the weighted sum of the Stability functions $R_{b_a}(z)$ and $R_{b_b}(z)$

Proof:
\begin{multline}
R_{\chi b_a + (1-\chi) b_b}(z) = R(z) = 1 + (\chi b_a + (1-\chi) b_b)^Tz(I - zA)^{-1}e \\
= \chi + (1-\chi) + \chi b_a ^Tz(I - zA)^{-1}e + (1-\chi) b_b^Tz(I - zA)^{-1}e \\
= \chi R_{b_a}(z) + (1-\chi) R_{b_b}(z)
\end{multline}

@@

For most methods, a small change of $b$ does not alter the stability function dramatically.
(TODO: I am missing a piece here from ‘the derivative is finite’ to ‘is similar’)

@@

The stability function depending on the $b$ is expressed 
When adapting the $b$ we cannot change them independently but only in ways that comply with the order conditions.
$m$ is the number of dimensions of the Kernel of the Order Condition $m = \mathrm{dim}\{\mathrm{Ker} (Q) \}$
The $b_1,\cdots,b_m$  are the Basis vectors of $\mathrm{Ker} (Q)$. With an coordinate Transformation we get the coordinate vector $\hat{b}$ with the Matrix $B = [b_1,\cdots,b_m]$ and $b = B \hat{b}$.

@@

\begin{equation}\label{eq:gen_stabilityf}
R(\hat{b},z) = 1 +  (b_{orig} +B \hat{b})^T z(I - zA)^{-1}
\end{equation}
The border is defined by the implicit function 
\begin{equation}\label{eq:border}
|R(\hat{b},u+iv)|^2 -1 = 0
\end{equation}
This function can be viewed either as $\R^m \times  \CN \rightarrow \R$ or $\R^m \times  \R^2 \rightarrow \R$.
We are only concerned with the expansion or contraction. Because of this we are only interested in the change perpendicular to the border of the stability region.

@@

The derivative in respect to $\hat{b}_j$ is 
$ \frac{\partial f}{\partial \hat{b}_j}(\hat{b},a(\hat{b})) \Big|_{\hat{b}=0}$
this simplifies to 
\begin{multline}\label{eq:derivative_to_b}
 \frac{\partial f}{\partial \hat{b}_j}(\hat{b},a(\hat{b})) \Big|_{\hat{b}=0} = 
 \frac{\partial }{\partial \hat{b}}(R(\hat{b},z)R(\hat{b},z^*)) \\
 = (B\hat{b}_j)^T (I-zA)^{-1} R_{b_{orig}}(z^*) + (B\hat{b}_j)^T (I-z^*A)^{-1} R_{b_{orig}}(z)
\end{multline}
using the fact that $z_0$ is on the border of the stability region.
The derivative $(B\hat{b})^T (I-zA)^{-1} R_{b_{orig}}(z^*) (B\hat{b})^T (I-z^*A)^{-1} R_{b_{orig}}(z)$ is known to be finite for explicit methods, because it is a polynomial and $|z_0| < \infty$
The same process can be done for all points on the border of the stability region. 
As long as $\nabla|R(\hat{b},u+iv)|^2 \neq 0 \forall \hat{b} < \epsilon  \forall z$ the derivative $\frac{\partial a}{\partial \hat{b}_j} (\hat{b})$ is finite. This means that a small change the $b$ only leads to a small change of the stability region.

@@

\begin{align}
 u_i^{n+1} &\geq 0   \;   \forall {i \in \{1, \cdots,m \}}  \\
 u_i^n + h \sum_{j=0}^{s-1} b_j f_i^j  &\geq 0   \;   \forall {i \in \{1,\cdots,m \}}  
\end{align}
has to be fulfilled.
This infers $m$ positivity constraints to the optimization problem. These can be written as

\begin{equation}
u_i + \dt F  b \geq 0     
\end{equation}
where $F = \big[f^1 , \cdots f^{s-1}\big]$.

This causes unnecessary many inequality constraints on the LP-Problem. To simplify the problem, one want to reduce the number of positivity constraints by ignoring it for ODEs where there is no issue with positivity.
The new positivity constraints can be formulated as
\begin{equation}
u_i^n + \Delta b \sum_{j=0}^{s-1} b_j f_i^j  \geq 0   \;   \forall {i \in h \subseteq \{1,\cdot,m \}} 
\end{equation}
For this a set $h \subseteq \{1,\cdots,m \}$ that include all indecies that are needed for determining the $b$ has to be chosen.
